---
title: "PROJECT ME315"
output: html_document
date: "2023-06-28"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading files
```{r}
# Loading the events.csv file
events <- read.csv("events.csv")

# Loading the ginf.csv file
ginf <- read.csv("ginf.csv")

#library(factoextra)
library(cluster)
library(knitr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(ggfortify)
```

#Left join ginf to events 
```{r}
events <- left_join(events, ginf[,c('id_odsp','league','season','country')], by = "id_odsp")
```

##Data cleaning process 
```{r}

events$assist_method <- as.factor(events$assist_method)
events$id_odsp <- as.factor(events$id_odsp)
events$league <- as.factor(events$league)
ginf$league <- as.factor(ginf$league)
ginf$season <- as.factor(ginf$season)
events$is_goal <- as.logical(events$is_goal)
events$bodypart <- as.factor(events$bodypart)


levels(events$league) <- c('Bundesliga','Premier League','Ligue 1','Serie A','La Liga')
levels(ginf$league) <- c('Bundesliga','Premier League','Ligue 1','Serie A','La Liga')

```



###OVERVIEW 
```{r}
ginf$goals <- (ginf$fthg + ginf$ftag)
goals <- ginf %>% group_by(league) %>% summarise(goals = sum(goals))
total_matches <- ginf %>% count(league)

ratio <- left_join(total_matches,goals, by = 'league')
names(ratio)[2] = 'total_matches'
ratio$goals_per_match <- ratio$goals / ratio$total_matches

ratio <- mutate_each(ratio,funs(round(., 2)),-league) #rounding to 2 dp
ratio <- arrange(ratio,desc(goals_per_match))


#Ratio
sorted_ratio <- ratio[order(-ratio$goals), ]

#Presenting into a table
kable(sorted_ratio[, c("league","goals", "total_matches", "goals_per_match")], 
      col.names = c("League Name", "Goals", "Number of Matches", "Goals per Match"),
      align = "c")

```



#There is a correlation with the number of goals you score and the matches you play. 
```{r}
# Creating the scatter plot
scatter_plot <- ggplot(data = ratio, aes(x = goals, y = total_matches, color = league, size = goals_per_match)) +
  geom_point() +
  labs(x = "Number of Goals", y = "Number of Matches", title = "Goals vs Matches by League") +
  scale_color_discrete(name = "League") +  # Add legend title
  scale_size_continuous(range = c(3, 8)) +
  theme_minimal()


scatter_plot

```




### We see how intense/rough the top 5 leagues are. 
```{r}

cards_shown <- events[events$event_type %in% c(4, 5, 6),] %>%
  group_by(league) %>%
  summarize(total = n())  
cards_shown$cards_per_match <- round(cards_shown$total / total_matches$n, 2)
cards_shown <- arrange(cards_shown, cards_per_match) 
cards_shown$league <- factor(cards_shown$league, levels = cards_shown$league[order(cards_shown$cards_per_match)]) 


my_palette <- c("tomato2", "orange", "yellow", "green", "blue")

ggplot(cards_shown, aes(x = league, y = cards_per_match, fill = cards_per_match)) +
  geom_bar(stat = "identity", width = 0.6) +
  scale_fill_gradient(low = my_palette[1], high = my_palette[length(my_palette)], guide = "legend") +
  labs(title = "Cards per Match for each League (Yellow + Red)",
       x = "League", y = "Cards Per Match") +
  theme(plot.title = element_text(size = 15, hjust = -0.6),
        axis.title = element_text(size = 12)) +
  geom_text(aes(label = cards_per_match), hjust = 2, size = 3, color = 'white') +
  coord_flip()


```


```{r}


team_goals <- events %>%
  group_by(event_team) %>%
  summarise(goals = sum(is_goal == TRUE)) %>%
  arrange(desc(goals))

team_goals


```

### Teams categorised by leagues (Cards per match)
```{r}
cards_shown <- events[events$event_type %in% c(4, 5, 6), ] %>%
  group_by(league, event_team) %>%
  summarize(total = n())  

total_matches <- events %>%
  group_by(league) %>%
  summarize(n = n_distinct(id_odsp))

cards_shown <- left_join(cards_shown, total_matches, by = "league")

cards_shown$cards_per_match <- round(cards_shown$total / cards_shown$n, 2)

cards_shown
cards_avg <- cards_shown %>%
  group_by(league) %>%
  top_n(5, cards_per_match) %>%
  ungroup()

ggplot(cards_avg, aes(x = reorder(event_team, cards_per_match), y = cards_per_match, fill = league)) +
  geom_bar(stat = "identity", width = 0.6) +
  labs(title = "Top 5 Teams: Cards per Match by League", x = "Team", y = "Cards Per Match") +
  theme(plot.title = element_text(size = 15, hjust = -0.6), axis.title = element_text(size = 12)) +
  geom_text(aes(label = cards_per_match), hjust = 0.5, vjust = -0.5, size = 3, color = "white") +
  coord_flip() +
  scale_fill_brewer(palette = "Set1")



```
### Hitting the post 

#Shows that they have taken a lot of shots and chances BUT unlucky :( 
```{r}
bars <- events[events$shot_outcome==4, ] %>%
  group_by(event_team,league) %>%
  summarize(total = n())
bars <- bars[with(bars,order(league,-total)),] #sort by league and bars
barsLeagueTeams <- by(bars, bars$league, function(X) X[1:4,])
bars <- do.call("rbind", barsLeagueTeams) #get top 4 teams in each league
bars$event_team <- factor(bars$event_team, levels = bars$event_team[order(bars$total)]) #reorder for plot

ggplot(bars, aes(x = event_team, y = total, fill = league)) +
  geom_bar(stat = "identity", width = 0.8, color = "black") +
  labs(title = "Most shots against the goal post (hitting the bar)",
       x = 'Team', y = 'Total Shots') +
  geom_text(aes(label = total), vjust = -0.5, size = 3, color = 'black') +
  theme(plot.title = element_text(size = 17, hjust = -0.7),
        legend.position = 'bottom') +
  facet_wrap(~league, nrow = 5, scales = 'free', strip.position = 'right') +
  coord_flip()

```



```{r}
top_goalscorers <- events[events$is_goal, ] %>%
  group_by(event_team, bodypart) %>%
  summarize(goals = n()) %>%
  arrange(desc(goals))

levels(top_goalscorers$bodypart) <- c('Right Foot', 'Left Foot', 'Header')
top_goalscorers <- top_goalscorers[with(top_goalscorers, order(bodypart, -goals)), ] # Sort by method and team
top_goalscorers_method <- by(top_goalscorers, top_goalscorers$bodypart, function(X) X[1:5, ])
top_goalscorers <- do.call("rbind", top_goalscorers_method) # Get top 5 teams in each category

top_goalscorers

ggplot(top_goalscorers, aes(x = event_team, y = goals, fill = bodypart)) +
  geom_bar(stat = "identity", width = 0.6, color = "black") +
  labs(title = "Methods of top goal-scoring teams",
       x = "Team", y = "Total Goals") +
  geom_text(aes(label = goals), vjust = -0.5, size = 3, color = "black") +
  theme(plot.title = element_text(size = 17, hjust = -0.7),
        legend.position = "bottom") +
  facet_wrap(~ bodypart, scales = "free") +
  ylim(0, max(top_goalscorers$goals)) +
  coord_flip()
```


```{r}

library(viridis)

#first 15 minutes 
first_15 <- events[events$time <= 15, ]

##last 15 minutes 
last_15 <- events[events$time >= 75 & events$time <= 90, ]

## Grouping by teams 
top_10_scorer_first_15 <- first_15[first_15$is_goal == 1, ] %>%
  group_by(event_team) %>%
  summarize(goals = n()) %>%
  arrange(desc(goals)) %>%
  top_n(10)


teams <- top_10_scorer_first_15$event_team

scores <- top_10_scorer_first_15$goals

## Plotting 
ggplot(top_10_scorer_first_15, aes(x = reorder(event_team, goals), y = goals, fill = goals)) +
  geom_bar(stat = "identity", width = 0.5) +
  scale_fill_viridis(option = "A", direction = -1) +
  labs(x = "Teams", y = "# of goals", title = "Most goals in the 1st 15 minutes") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        panel.grid.major.y = element_blank(),
        panel.grid.minor.y = element_blank(),
        legend.position = "none") +
  coord_flip()

#----------------------------------------------------------------------------------------------------------------
## Grouping by teams 
top_10_scorer_last_15 <- last_15[last_15$is_goal == 1, ] %>%
  group_by(event_team) %>%
  summarize(goals = n()) %>%
  arrange(desc(goals)) %>%
  top_n(10)


teams_last_15 <- top_10_scorer_last_15$event_team

scores_last_15 <- top_10_scorer_last_15$goals

## Plotting
ggplot(top_10_scorer_last_15, aes(x = reorder(event_team, goals), y = goals, fill = goals)) +
  geom_bar(stat = "identity", width = 0.5) +
  scale_fill_viridis(option = "A", direction = -1) +
  labs(x = "Teams", y = "Number of goals", title = "Most goals in the last 15 minutes") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        panel.grid.major.y = element_blank(),
        panel.grid.minor.y = element_blank(),
        legend.position = "none") +
  coord_flip()

```

### Extracting all the factors above 
```{r}


team_goals <- events %>%
  group_by(event_team) %>%
  summarize(goals = sum(is_goal == TRUE))


team_cards <- events %>%
  filter(event_type %in% c(4, 5, 6)) %>%
  group_by(event_team) %>%
  summarize(cards = n()) 

team_post_hits <- events %>%
  filter(shot_outcome == 4) %>%
  group_by(event_team) %>%
  summarize(post_hits = n()) %>%
  ungroup() %>%
  right_join(distinct(events, event_team), by = "event_team") %>%
  mutate(post_hits = replace_na(post_hits, 0))

team_goal_method <- events %>%
  filter(is_goal == 1) %>%
  group_by(event_team, bodypart) %>%
  summarize(method_goals = n()) %>%
  spread(bodypart, method_goals, fill = 0) %>%
  ungroup()

team_goal_time_first15 <- events %>%
  filter(is_goal == 1, events$time <= 15) %>%
  group_by(event_team) %>%
  summarize(total_goals = n()) %>%
  ungroup()

team_goal_time_last15 <- events %>%
  filter(time >= 75 & time <= 90) %>%
  group_by(event_team) %>%
  summarize(total_goals = sum(is_goal == 1)) %>%
  ungroup() %>%
  right_join(distinct(events, event_team), by = "event_team") %>%
  mutate(total_goals = replace_na(total_goals, 0))

full_data <- data.frame(
  team_goals$goals,
  team_cards$cards,
  team_post_hits$post_hits,
  as.matrix(team_goal_method[, -1]),  # Exclude the event_team column
  team_goal_time_first15$total_goals,
  team_goal_time_last15$total_goals
)


```




### Complete Linkage vs Single Linkage 
```{r}

# distance
distance_matrix <- dist(full_data)

# complete and single linkage
hierarchical_complete <- hclust(distance_matrix, method = "complete")
hierarchical_single <- hclust(distance_matrix, method = "single")

# choosing k 
num_clusters <- 3

# clustering
clusters_complete <- cutree(hierarchical_complete, k = num_clusters)
clusters_single <- cutree(hierarchical_single, k = num_clusters)

# cluster table
cluster_table_complete <- data.frame(event_team = team_goals$event_team, cluster = clusters_complete)
cluster_table_single <- data.frame(event_team = team_goals$event_team, cluster = clusters_single)

# labeling
labels <- team_goals$event_team


par(mfrow = c(1, 3))

# plotting + categorizing the clusters with label (complete)
plot(hierarchical_complete, main = "Complete Linkage", xlab = "Teams", ylab = "Distance", labels = labels, hang = -1, cex = 0.8)
rect.hclust(hierarchical_complete, k = num_clusters, border = num_clusters:1)


#plotting + categorizing the clusters with label (single)
plot(hierarchical_single, main = "Single Linkage", xlab = "Teams", ylab = "Distance", labels = labels, hang = -1, cex = 0.8)
rect.hclust(hierarchical_single, k = num_clusters, border = num_clusters:1)

# cluster tab
print(cluster_table_complete)
print(cluster_table_single)


```







#Complete Linkage is better so we do this again with another variable name
```{r}


hierarchical_cluster <- hclust(distance_matrix)


num_clusters <- 3
clusters <- cutree(hierarchical_cluster, k = num_clusters)


cluster_table <- data.frame(event_team = team_goals$event_team, cluster = clusters)


labels <- team_goals$event_team


plot(hierarchical_cluster, main = "Hierarchical Clustering Dendrogram", xlab = "Teams", ylab = "Distance", labels = labels, hang = -1, cex = 0.8)


rect.hclust(hierarchical_cluster, k = num_clusters, border = num_clusters:1)



```


#Printing Cluster Tables
```{r}
cluster_data <- data.frame(Cluster = c("Cluster 3", "Cluster 2", "Cluster 1"),
                           Teams = c(paste(cluster_table$event_team[cluster_table$cluster == 3], collapse = ", "),
                                     paste(cluster_table$event_team[cluster_table$cluster == 2], collapse = ", "),
                                     paste(cluster_table$event_team[cluster_table$cluster == 1], collapse = ", ")))



#Tables
kable(cluster_data, format = "markdown", align = c("l", "l"), col.names = c("Cluster", "Teams"))
```
###We can see that Barcelona  "Bayern Munich" "Borussia Dortmund" "Juventus""Lyon""Napoli"Paris Saint-Germain" "Real Madrid" are dominating in terms of their playstyle at the moment.




#Finding out optimum K 
```{r}
wssplot <- function(data, nc = 2:15, seed = 1) {
  wss <- numeric(length(nc))
  for (i in seq_along(nc)) {
    set.seed(seed)
    wss[i] <- sum(kmeans(data, centers = nc[i])$withinss)
  }
  plot(nc, wss, type = 'b', xlab = 'Number of Clusters', ylab = 'Within groups Sum of Squares')
}

data <- full_data  
wssplot(data)

```
 
 
```{r}
# k menas
KM <- kmeans(full_data, 4, nstart = 20)

# cluster assignment for each data pt
cluster_assignments <- KM$cluster

# data frame for plotting
plot_data <- cbind(full_data, Cluster = cluster_assignments)

# plotting
autoplot(KM, data = plot_data, frame = TRUE, label = TRUE, label.size = 3)

# cluster centers
print(KM$centers)

# table
cluster_table <- data.frame(Team = unique(events$event_team), Cluster = cluster_assignments)

# display table
print(cluster_table)

```


#Who are in cluster 2?  
```{r}
# filtering the cluster_table for teams in cluster 2
teams_in_cluster_2 <- cluster_table$Team[cluster_table$Cluster == 1]

# priting the teams
print(teams_in_cluster_2)
```

